---
title: "1 Getting started"
output: rmarkdown::html_vignette
knit: (function(inputFile, encoding, output_dir) { 
      rmarkdown::render(inputFile,
                        encoding=encoding, 
                        output_dir=paste0(dirname(normalizePath(getwd())), "/inst/doc"))})
vignette: >
  %\VignetteIndexEntry{1 Getting started}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

For installation and configuration, see 
[here](https://r-and-gama.github.io/gamar/index.html#installation-and-configuration).

```{r}
library(gamar)
```

## 1.1 Desiging experiments

We can make an `experiment` object either by reading an experiment from a GAML
file with the `load_experiment()` funtion or by building it from scratch with
the `experiment()` constructor.

### 1.1.1 Loading an experiment from a GAML file

Let's use the following `gaml` file:

```{r}
gaml_file <- system.file("models", "sir.gaml", package = "gamar")
```

Loading the experiment `sir` from this file. Note that we also indicate the
names of the folder where we wish to save the simulations outputs. If not
specified, the name of the `.gaml` is used by default:

```{r}
exp1 <- load_experiment("sir", gaml_file)
```

which gives:

```{r}
exp1
```

An `experiment` is a simple data frame:

```{r}
class(exp1)
```

This implies that we will be able to use all the data frame methods to
manipulate experiments. 

### 1.1.2 Building an experiment from scratch

Alternatively, an experiment can be built from scratch with the `experiment`
constructor:

```{r}
df <- data.frame("S0" = 999, "I0" = 1, "R0" = 0, "beta" = 1.5,
                 "gamma" = 0.15, "S" = 1, "I" = 1, "R" = 1,
                 "a" = 2, "b" = 1)
```


```{r}
exp2 <- as_experiment(df = df,
           parameters = c("S0", "I0", "R0", "beta", "gamma"),
           obsrates = c("S", "I", "R"), tmax = "a", seed = "b",
           experiment = "sir", model = gaml_file)
```

Which gives the same thing:

```{r}
exp2
```

Note that if you enter values of inequal lengths, it will automatically
complete:

```{r}
df <- data.frame("S0" = c(999, 990), "I0" = c(1, 2), "R0" = 0, "beta" = 1.5,
                 "gamma" = 5, "S" = 1, "I" = 1, "R" = 1,
                 "a" = 2, "b" = 1)
exp3 <- as_experiment(df = df, parameters = c(1:5),
           obsrates = c(6:8), tmax = "a", seed = "b",
           experiment = "sir", model = gaml_file)
```

Which gives:

```{r}
exp3
```

Note finally, that we can make smart use of the `expand.grid()` function of the
`base` package to efficiently generate a complete experimental design:

```{r}
df <- expand.grid(S0 = c(900, 950, 999),
               I0 = c(100, 50, 1),
               R0 = 0,
               beta = 1.5,
               gamma = .15,
               S = 1,
               I = 1,
               R = 1,
               tmax = 2,
               seed = 1)

exp4 <- as_experiment(df, parameters = c(1:5),
           obsrates = c(6:8), tmax = "tmax", seed = "seed",
           experiment = "sir", model = gaml_file)
```

Which gives:

```{r}
exp4
```

We'll see more about that in the next sections.

### 1.1.3 Building an experiment from a data frame

```{r}
df <- data.frame(
  S0 = c(700, 800, 900, 950, 999),
  R0 = 0,
  beta = 1.5,
  gamma = .15,
  S = 1,
  I = 1,
  tmax = 2,
  seed = sample(5)
)
```

Constant population size of 1000:

```{r}
df$I0 <- 1000 - df$S0
```

```{r}
df
```

Let's now transform this data frame into an `experiment`:

```{r}
exp5 <- as_experiment(df = df,
           parameters = c("S0", "R0", "I0", "beta", "gamma"),
           obsrates = c("S", "I"), tmax = "tmax", seed = "seed",
           experiment = "sir", model = gaml_file)
```

Which gives:

```{r}
exp5
```

## 1.2 Manipulating experiments

As mentioned above, since `experiment` objects are data frames, all the data
frame methods can be used to manipulate them, such as `nrow()` to know the
number of simulations in an experiment:

```{r}
nrow(exp4)
```


### 1.2.1 Replicating experiments

From the following experiment

```{r}
exp1
```

We can replicate:

```{r}
exp5 <- repl(exp1, 10)
```

Which gives:

```{r}
exp5
```

On which we could use different seeds for each simulation:

```{r}
exp5$seed <- sample(10)
```

Which gives:

```{r}
exp5
```


### 1.2.2 Extracting parameters values

```{r}
parameters(exp5)
```


### 1.2.3 Extracting observation rates

```{r}
obs_rates(exp5)
```


### 1.2.4 `tmax` and `seed`

As seen above, `seed` can be extracted and changed with regular data frame
methods:

```{r}
exp5$seed
exp5$seed <- sample(10)
exp5
```

Same for `tmax`:

```{r}
exp5$tmax
exp5$tmax <- 3L
exp5
```

If you want to stop observing the variable R and observe the variable I every
2 steps, you'd do:

```{r}
exp5$r_R <- NULL
exp5$r_I <- 2
```

Which gives:

```{r}
exp5
```


## 1.3 Running experiments

Then, we run the experiment by one command `run_experiment`, in which we specify
the experiment and optionally number of cores, output folder for the results of
this experiment, parameters file in xml format. 

```{r}
model(exp1)$md5sum
```

```{r}
output <- run_experiment(exp1, hpc = 2)
```

Which gives:
```{r}
output
```

```{r cleaning, include = FALSE}
unlink("sir*", TRUE)
```

